---
title: "Data Tools 2: A New Variable for Classifying Marital Status in the American Community Survey"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

An R reproduction of STATA code in [Data Tools 2: A New Variable for Classifying Marital Status in the American Community Survey](https://www.aei.org/research-products/working-paper/data-tools-2-a-new-variable-for-classifying-marital-status-in-the-american-community-survey/) by Charles Murray.

Running this file in R will convert the IPUMS source data in this directory to the form discussed in the article above.

### Load and install (if necessary) packages

```{r, echo=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman", repos = "https://cloud.r-project.org")
pacman::p_load(ipumsr, data.table, scales, testthat)
```

### Download data to folder
**Skip this section if you don't have a beta API key
```{r eval=FALSE}
# sample reference: https://usa.ipums.org/usa-action/samples/sample_ids
samples <- paste0('us', 2013:2020, 'a')

# define the data extract
extract <- define_extract_usa(description = "Replication of Charles Murray's Data Tools part 2", samples =  samples,
  variables = c('YEAR', 'SERIAL', 'PERNUM', 'RELATED', 'SPLOC', 'MARST',
    'SEX'),
  data_format = 'fixed_width')

# submit the data request to IPUMS
submitted_extract <- submit_extract(extract)

# save the extract for reproduction
save_extract_as_json(extract, 'ipums_extract.json')
```

##### Have R check periodically until the extract is ready
```{r eval=FALSE}
extract <- define_extract_from_json('ipums_extract.json')

submitted_extract <- submit_extract(extract)
# will print TRUE when ready, may take a few minutes
wait_for_extract(submitted_extract)

workingset_path <- download_extract(submitted_extract)
```

*Alternatively, if you download the extract directly from IPUMS, you can use the below:*
```{r}
workingset_path <- 'usa_00021.xml' # your number will vary.
```


### Follow the instructions in the article

1. Open the dataset you want to analyze. In this example, I’ll call it workingset. Sort it by YEAR, SERIAL, and PERNUM.
```{r}
ddi <- read_ipums_ddi(workingset_path)
workingset <- read_ipums_micro(ddi) 

# set to data.table for speed
setDT(workingset, key = c('SERIAL', 'PERNUM'))

setorder(workingset, YEAR, SERIAL, PERNUM)
```

2. Keep YEAR, SERIAL, PERNUM, RELATED, SPLOC, MARST, and SEX, and drop the rest of the variables. Save this temporarily reduced database with its own name—say,
temp.
```{r}
temp <- copy(workingset[, list(YEAR, SERIAL, PERNUM, RELATED, SPLOC, MARST, SEX)])
```

3. Drop all the cases of people who do not have a partner living in the household
(SPLOC = 0).

```{r}
temp <- temp[SPLOC != 0]
```

4. Drop PERNUM and then rename SPLOC as PERNUM.

```{r}
temp$PERNUM <- NULL
setnames(temp, 'SPLOC', 'PERNUM')
```

5. Rename RELATED, MARST, and SEX so they reflect the values associated with the
partner—for example, PARTRELATED, PARTMARST, and PARTSEX.

```{r}
setnames(temp, c('RELATED', 'MARST', 'SEX'), 
  c('PARTRELATED', 'PARTMARST',  'PARTSEX'))
```

6. Sort YEAR, SERIAL, and PERNUM. Save the revised temp.
```{r}
setorder(temp, YEAR, SERIAL, PERNUM)
```

**Not necessary** 7. Reopen workingset.

8. Merge workingset with temp using YEAR, SERIAL, and PERNUM
```{r}
workingset <- merge(workingset, temp, by = c('YEAR', 'SERIAL', 'PERNUM'),
  all.x = T)
```

This reproduces the dataset used for analysis:
```{r}
workingset
```

## Coding ALTMARST

The codes for MARST are:
```{r}
ddi$var_info$val_labels[[8]]
```

The RELATED code used to create ALTMARST is 1114 (“Unmarried partner”).
The codes for SEX are 1 for male and 2 for female.

Here is the code for creating ALTMARST

```{r}
# cast the variables to lower case
setnames(workingset, tolower)
```

#### MARRIED
```{r}
workingset[, altmarst := fcase(
# married, spouse present, straight
  sploc>0 & marst==1 & partmarst==1 & sex!=partsex, 10,
  # married, spouse present, lesbian
  sploc>0 & marst==1 & partmarst==1 & sex==2 & partsex==2, 11,
  # married, spouse present, gay
  sploc>0 & marst==1 & partmarst==1 & sex==1 & partsex==1, 12,
  # married, spouse absent
  sploc==0 & marst==2, 13
)]
```


#### COHABITING
```{r}
workingset[is.na(altmarst), altmarst := fcase(
  # cohabiting, straight
  related==1114 & sex!=partsex, 20,
  partrelated==1114 & sex!=partsex, 20,
  # cohabiting,lesbian
  related==1114 & sex==2 & partsex==2, 21,
  partrelated==1114 & sex==2 & partsex==2, 21,
  # cohabiting, gay
  related==1114 & sex==1 & partsex==1, 22,
  partrelated==1114 & sex==1 & partsex==1, 22
)]
```



#### LIVING ALONE
```{r}
workingset[is.na(altmarst), altmarst := fcase(
  # single, separated
  sploc==0 & marst==3 & related!=1114, 30,
  # single, divorced
  sploc==0 & marst==4 & related!=1114, 31,
  # single, widowed
  sploc==0 & marst==5 & related!=1114, 32,
  # single, never-married
  sploc==0 & marst==6 & related!=1114, 33
)]
```


#### CREATE ALTMARST VALUE LABELS

**Using R's factors are the closest in concept to STATA labels**

```{r}
workingset[, altmarst_factor := factor(altmarst, 
  levels = as.character(c(10:13, 20:22, 30:33)),
  labels = c("Married, straight", "Married, lesbian",
 "Married, gay", "Married, spouse absent", "Cohabit, straight",
 "Cohabit, lesbian", "Cohabit, gay", "Single, separated",
 "Single, divorced", "Single, widowed", "Single, never married")
)]

workingset[, list(.N), keyby = c('altmarst', 'altmarst_factor')]
```


## Replicating the results
```{r}
mar_lesbian <- (sum(workingset$altmarst == 11))
mar_gay <- (sum(workingset$altmarst == 12))
cohab_lesbian <- (sum(workingset$altmarst == 21))
cohab_gay <- (sum(workingset$altmarst == 22))

#test
expect_equal(mar_lesbian, 48846)
expect_equal(mar_gay, 47604)
expect_equal(cohab_lesbian, 35424)
expect_equal(cohab_gay, 35536)

```


First, ALTMARST enables researchers to study the characteristics of same-sex couples. Combining the ACS surveys from 2013–20 produces samples of **`r comma(mar_lesbian)`** married lesbian couples, **`r comma(mar_gay)`** married gay couples, **`r comma(cohab_lesbian)`** cohabiting lesbian couples, and **`r comma(cohab_gay)`** cohabiting gay couples—samples large enough to permit many useful analyses.